---
title: "Lytras RHAS HMM"
author: "Fred Jaya"
output: 
   html_document:
     theme: spacelab
     code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T,
  warning = F, 
  dev = 'png',
  fig.width = 8,
  fig.path = "/home/fredjaya/GitHub/rec-mast/lytras/figs/")

library(MixtureModelHMM)
library(ggplot2)
library(dplyr)
library(tidyr)
library(flextable)
```

## Initial single tree 

```{r data_prep}
site_prob <- read.table("rates_i0.siteprob", header = T)
aln_info <- read.table("rates_i0.alninfo", header = T)  

sites <- 
  site_prob %>%
  left_join(aln_info, by = "Site") %>%
  mutate(Stat = gsub("-", "All gaps", Stat)) %>%
  mutate(Stat = gsub("C", "Constant", Stat)) %>%
  mutate(Stat = gsub("c", "Constant+ambiguous", Stat)) %>%
  mutate(Stat = gsub("I", "Informative", Stat)) %>%
  mutate(Stat = gsub("U", "Uninformative", Stat))

rm(site_prob, aln_info)

sites_long <-
  sites %>%
  pivot_longer(p1:p4, names_to = "Rate class", values_to = "Posterior probability")
```

### Types of sites  
```{r site_type_freq}
n_sites <- nrow(sites)

sites %>%
  # Count number of site types
  select(Stat) %>%
  table() %>%
  as.data.frame() %>%
  rename("Site type" = '.', "Frequency" = Freq) %>%
  # Add proportions
  mutate(Proportion = round(Frequency / n_sites, 5)) %>%
  flextable()
```


### Site probabilities  
```{r singletree_siteprob_allsites}
sites_long %>%  
  ggplot(aes(x = Site, y = `Posterior probability`, colour = `Rate class`)) +
  geom_point(size = 0.01, alpha = 0.1) +
  geom_smooth(method = "gam", formula = y~s(x, bs = 'cs', fx = T, k = 50)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_light() +
  labs(title = "All sites", subtitle = "n = 30,956")
  
#plot_scatter("rates_i0.siteprob") +
#  labs(y = "Posterior probability")

#plot_scatter("rates_i0.sitelh") +
#  labs(y = "lnL")
```

```{r singletree_siteprob_infsites}
sites_long %>%
  filter(Stat == "Informative") %>%
  ggplot(aes(x = Site, y = `Posterior probability`, colour = `Rate class`)) +
  geom_point(size = 0.01, alpha = 0.2) +
  geom_smooth(method = "gam", formula = y~s(x, bs = 'cs', fx = T, k = 50)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +  
  theme_light() +
  labs(title = "Informative sites only", subtitle = "n = 13,321")
```

Class `p1` appears to be dominated by constant or uninformative sites. When only informative sites are retained, `p2` becomes the prevalent class. In fact, the `p1` kernel follows the same trend as the dominant `p2` class. There is a clear transition between classes `p2` to `p3`/`p4` at positions ~22k and 28k, and less so around ~4k. Class `p4` appears to have a low posterior probability across sites, until sites 22~28k.  

This data is suitable for analysis with a HMM.  

## Running the HMM

```{r hmm_prep}
hmm_prob <- run_HMM(site_info = "rates_i0.siteprob", "rates_i0.alninfo")
hmm_lnl <- run_HMM(site_info = "rates_i0.sitelh", "rates_i0.alninfo")
```

### HMM summaries

```{r hmm_summary}
summary(hmm_prob)
summary(hmm_lnl)
```
Posterior probabilities and site likelihoods looks identical. Proceed with site likelihoods.

```{r hmm_alnplot_lnl, fig.height=2}
hmm_lnl$alignment_plot +
  theme_light()
```
```{r hmm_lnl_stats}
#hmm_lnl$hmm_transition_table
#hmm_lnl$classification
```
Output partitions.  
```{r}
#save_partitioning_scheme(hmm_lnl, "rates_i0_rates.nex")
```
# MAST +T

```{r}
plot_scatter("rates_i0_mast_T.siteprob") +
  labs(y = "Posterior probability")

plot_scatter("rates_i0_mast_T.sitelh") +
  labs(y = "lnL")
```

```{r}
hmm_lnl_mast <- run_HMM(site_info = "rates_i0_mast_T.sitelh", "rates_i0_mast_T.alninfo")
hmm_lnl_mast$alignment_plot
```
